---
title: 'Chapter 3: where to go next'
author: "Harro Cyranka"
date: "1/4/2019"
output: html_document
---
##Loading libraries
```{r, echo=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(gapminder)
library(here)
```

##Where to go next 1
What happens when geom_smooth() precedes geom_point(). The points are drawn OVER the smooth. This is useful when building plots, because we can work in layers, adding new information as we progress.

```{r, echo=FALSE, warning=FALSE, message = FALSE}
gapminder:: gapminder %>%
    ggplot(aes(x = gdpPercap, y= lifeExp)) + 
    geom_smooth() + geom_point()
```

##Where to go next 2
Replacing per capita GDP with population. Each point represents a country in a different year.
```{r, echo=FALSE, warning = FALSE, message = FALSE}
gapminder:: gapminder %>%
    ggplot(aes(x = pop, y = lifeExp)) + 
    geom_point() + geom_smooth()

```


##Where to go next 4
Mapping color to year instead of continent. Since year is a continuous variable, the points get shaded using a gradient scale.

```{r, echo=FALSE, warning = FALSE, message = FALSE}
gapminder::gapminder %>%
    ggplot(aes(x = gdpPercap, y = lifeExp, color = year)) + 
    geom_point() + scale_x_log10()

```


##Where to go next 5
Changing year to a factor variable and mapping it to color. Now each year gets shaded with a color from the palette.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
gapminder::gapminder %>%
    ggplot(aes(x = gdpPercap, y = lifeExp, color = factor(year))) + 
    geom_point() + scale_x_log10()

```


##Where to go next 6
Adding some structure to the data.

```{r, echo=FALSE, warning = FALSE, message = FALSE}
gapminder::gapminder %>%
    ggplot(aes(x = gdpPercap, y = lifeExp)) + 
    geom_point() + geom_smooth() + 
    facet_wrap(~year) + 
    scale_x_log10()
    
```