---
title: 'Chapter 4: where to go next'
author: "Harro Cyranka"
date: "1/7/2019"
output: html_document
---

##Loading libraries
```{r, echo=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(gapminder)
library(here)
library(socviz)
```


1) Using different ways to facet the data, using the gapminder data.
1.1 Faceting data on year
```{r, echo = FALSE, message=FALSE, warning=FALSE}
p <- ggplot(data = gapminder, aes(x = gdpPercap,
                                  y = pop))
p + geom_point(aes(group = country)) + 
    facet_wrap(~year)
```

1.2 Faceting data on country
```{r, echo=FALSE, message=FALSE, warning=FALSE}
p <- ggplot(data = gapminder, aes(x = gdpPercap,
                                  y = pop)) + 
    geom_point(aes(group = country)) + 
    facet_wrap(~country)

ggsave("faceting_data_by_country.png", plot = p, dpi = 400)
```

2) Using GSS data to differentiate between two calls of facet_grid
2.1 facet_grid(sex ~ race): gender are rows, while race are columns
```{r, echo = FALSE, warning = FALSE, message=FALSE}
p <- ggplot(data = gss_sm,
            mapping = aes(x = age, y = childs))
p + geom_point(alpha = 0.2) + 
    geom_smooth() + 
    facet_grid(sex~race) ##Wall against ground

```

2.2 facet_grid(race ~ sex): race are rows, while gender are columns


```{r, echo = FALSE, warning = FALSE, message=FALSE}
p <- ggplot(data = gss_sm,
            mapping = aes(x = age, y = childs))
p + geom_point(alpha = 0.2) + 
    geom_smooth() + 
    facet_grid(race~sex) ##Wall against ground

```


3) Using more complex formulas with facet_wrap.
Graphs are equal to facet_grid(sex~race). However it does not wrap results in a fully cross-classified grid.
```{r, echo = FALSE, warning = FALSE, message=FALSE}
p <- ggplot(data = gss_sm,
            mapping = aes(x = age, y = childs))
p + geom_point(alpha = 0.2) + 
    geom_smooth() + 
    facet_wrap(~sex + race) ##Wall against ground

```

4) Using geom_freqpoly instead of histograms
```{r, echo = FALSE, warning = FALSE, message=FALSE}
p <- ggplot(data = midwest,
            mapping = aes(x = poptotal))
p + geom_freqpoly() + 
    scale_x_log10()
```

5) Using geom_bin2d()
* These are useful to show densities for two variables at the same time
```{r, echo=FALSE, warning=FALSE, message=FALSE}
p <- ggplot(data = gapminder, 
            mapping = aes(x =gdpPercap,
                          y = lifeExp))

p + geom_bin2d(bins= c(20,50))
```

* Using geom_hexagon, which is similar
```{r, echo=FALSE, warning=FALSE, message=FALSE}
p <- ggplot(data = gapminder, 
            mapping = aes(x =gdpPercap,
                          y = lifeExp))

p + geom_hex()
```

6) Using geom_density_2d

*These graphs are similar to contour plots
6.1: Using without a geom_point() layer
```{r, echo=FALSE, warning=FALSE, message=FALSE}
p <- ggplot(data = midwest, aes(x = percbelowpoverty,
                                y = percollege))
p + geom_density_2d()
```

6.2: Using with a geom_point() layer
```{r, echo=FALSE, warning=FALSE, message=FALSE}
p <- ggplot(data = midwest, aes(x = percbelowpoverty,
                                y = percollege))
p + geom_point(size = 4, alpha = 0.5) + 
    geom_density_2d()
```